Evaluating contact tracing and isolation to control pandemic potential influenza outbreaks

### 2025-04-07 inst/paper/main.tex
update paper

The likelihood of an outbreak being controlled, in other words going locally-extinct or -eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical. The primary epidemiological determinant for the spread of an infectious disease is the reproduction number, defined as the average number of secondary cases  per primary case. The higher the reproduction number the less likely a disease, without intervention, is to cease transmission in a population. If a pathogen is subcritical (i.e. R $<$ 1) it will eventually stop spreading but can still cause transient outbreaks (Farrington and Grant, 1999). Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process (Lloyd-Smith et al., 2005), with higher individual-level variation for a given reproduction leading to more outbreaks being controlled (however other outbreaks may become larger due to superspreading). In addition, the likelihood of containment is reduced (assuming R $>$ 1) if there are multiple introductions into a populations (Kucharski et al., 2020). \\

@@

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts (ref; Kucharski et al., 2016). Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs). However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely (Dhillon and Srikrishna, 2018; ref).

@@

- Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
- SI shorter for pandemic flu (ref).

@@

The novelty of a pathogen introduced into a human population, in the sense of antigenically novel, is a leading risk factor for causing a pandemic (ref). Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 and H7N9 (Tanner et al., 2015) and was the case when H1N1 caused an outbreak in 2009 (ref). \\

@@

In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential. We include A/H5N1, A/H1N1, and A/H7N9. These have caused outbreaks of various size and severity over the past 100 years. H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission (Yang et al., 2007; Aditama et al., 2012). H1N1 (Fraser et al. 2009). \\

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../plots/prop_presymptomatic_transmission.png}
  \caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skewed-normal distribution.}
  \label{fig:prop-presym_trans}
\end{figure}

@@

@@

>>>>

The likelihood of an outbreak being controlled, in other words going locally-extinct or -eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical. The primary epidemiological determinant for the spread of an infectious disease is the reproduction number, defined as the average number of secondary cases  per primary case. The higher the reproduction number the less likely a disease, without intervention, is to cease transmission in a population. If a pathogen is subcritical (i.e. $R < 1$) it will eventually stop spreading but can still cause transient outbreaks (Farrington and Grant, 1999). Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process (Lloyd-Smith et al., 2005). Higher individual-level variation for a given reproduction leading to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading). In addition, the likelihood of containment is reduced (assuming $R > 1$) if there are multiple introductions into a populations (Kucharski et al., 2020). \\

@@

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts (ref; Kucharski et al., 2016). Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs). However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely (Dhillon and Srikrishna, 2018; ref). \\

@@

- Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
- SI shorter for pandemic flu (ref). \\

@@

The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic (ref). Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 and H7N9 (Tanner et al., 2015) and was the case when H1N1 caused an outbreak in 2009 (ref). \\

@@

In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential. We include A/H5N1, A/H1N1, and A/H7N9. These have caused outbreaks of various size and severity over the past 100 years. H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission (Yang et al., 2007; Aditama et al., 2012). H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak (Fraser et al. 2009; Lessler et al., 2009; ref on history of H1N1pdm). \\

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../plots/prop_presymptomatic_transmission.png}
  \caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skewed-normal distribution.}
  \label{fig:prop-presym-trans}
\end{figure}

@@

We simulated scenarios with either 5 or 10 initial cases.

@@

Predictably, a higher reproduction number for the outbreak results in a smaller percentage of outbreaks controlled at a given percentage of control effectiveness. (Figure \ref{fig:prop-outbreak-control-R}). The disparity of percentage outbreaks controlled becomes zero as the percentage of contacts traced exceeds 80\% (Figure \ref{fig:prop-outbreak-control-R}). When $R$ is only slightly above unity but below 2, the percentage of outbreaks controlled at 50\% of contacts traced and isolated is over 90\%, highlighting the effectiveness of this intervention for infectious diseases are not highly transmissible between humans.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../plots/prop_outbreak_control_reproduction_number.png}
  \caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an assumed reproduction number ($R$) of either 1.1 or 1.5. This is for a baselines scenario where the number of initial cases is 10, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
  \label{fig:prop-outbreak-control-R}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../plots/prop_outbreak_control_num_init_cases.png}
  \caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an initial number of seeding infections assumed to be either 5 or 10. This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
  \label{fig:prop-outbreak-control-num-init-cases}
\end{figure}
